Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression

### 2018-12-20 paper/paper.tex
Renamed paper file to paper.tex. Final version submitted to arXiv.

@@

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817. We present a completed version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}. This model includes the quadrupolar and octopolar adiabatic and dynamical tides, and we improve it by including the spin-induced quadrupole moment effect and a prescription for tapering the end of the waveform to agree with numerical simulations. The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. 
%This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. 
Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations. We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression. The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz. With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can accurately measure quantities such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be stacked to significantly improve measurements of NS radii and the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.

The reliability of these measurements, however, depend on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data. Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. Depending on the model, this can take seconds to hours for long BNS waveforms, so analytic frequency-domain approximations are often used instead.

For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

In this paper, we will use an implementation of the effective one body (EOB) formalism that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note an alternative model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a slightly different point-particle and spin prescription. For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.

Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible. In Ref.~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. Unfortunately, direct comparisons between this code and traditional samplers have not been published for BNS systems using the same waveform models. It is therefore not clear how accurate this code is at measuring quantities such as the tidal parameters. As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for rapidly evaluating waveform models. The essential idea is to construct a fitting function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions.
The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
In this work, instead of using a reduced basis, we will instead use the fact that the Fourier transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we can efficiently represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

@@

This work also takes a different approach from the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.

%This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.

We organize the paper as follows. In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In Section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section IV. In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817. Finally, we discuss future improvements in Section VI. We also give the expressions for the TaylorF2 base model in the Appendix.

@@

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. We adopt the tidally-augmented expression for $\Delta_u$ discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}:
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass EOB term (Eq.~(2.2) of Ref.~\cite{Bohe:2016gbl}), and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. If either component of the binary is a black hole, then we set the tidal polarizabilities to zero. The tidal contribution, including quadrupolar and octupolar dynamical tides, is
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,.
\end{align}
\end{widetext}
Here, $M_{A,B}$ are the masses of bodies $A$ and $B$, $M = M_A + M_B$ is the total mass, $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$. Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor (Eq.~(11) of Ref.~\cite{Dietrich:2017feu}), which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$. This enhancement factor is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.

Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
\begin{equation}
h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
\end{equation} 
where the point-mass piece $h_{\ell m}^{\textrm{pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and the tidal piece $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}. The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the gravitational-wave flux (Eqs.~(5) and~(6) of Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived. The Hamiltonian equations of motion with this radiation-reaction force are numerically integrated beginning with the quasicircular initial conditions described in Ref.~\cite{Buonanno:2005xu}. The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics. We only include the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes when evaluating the radiation reaction force.

%Let $\Omega$ be the instantaneous orbital frequency, $r$ be orbital separation, $p_r$ be the radial momentum, $f_s$ be the waveform sampling frequency in Hz, $\omega_{\textrm{peak}}^{\textrm{NR}}$ be the NR fit for the (2,2)-mode amplitude peak of a BNS merger defined in Eq.~(2) of Ref.~\cite{Bernuzzi:2015rla}~\footnote{Note that this fit was tuned to NR simulations only for $0 \leq \kappa_2^T \leq 500$, where $\kappa_2^T$ is defined by Eq.~(1) of Ref.~\cite{Bernuzzi:2015rla}. In our model, for $\kappa_2^T >500$ we use the value of this fit at $\kappa_2^T=500$.}. Then the numerical integration of the orbital dynamics stops when any of the following events happens: (i) $\Omega$ peaks, (ii) $p_r$ becomes positive, (iii) $\dot{r}$ becomes positive, (iv) $\dot{p}_r$ becomes positive, (v) $r$ does not evolve more than a 1000th of the leading-order estimate for $\dot{r}$, that is $\propto 64/5 \nu/r^4$, (vi) $2\Omega$ reaches $\omega_{\textrm{peak}}^{\textrm{NR}}$, (vii) $2M\Omega$ reaches $\pi M f_s$. In order to avoid triggering these stopping conditions at low frequency, where residual orbital eccentricity due to imperfect quasicircular initial conditions could be present, we exploit the $\omega_{\textrm{peak}}^{\textrm{NR}}$ fit to define a radius below which the stopping conditions can be triggered; such threshold value is taken to be $1.5 (\omega_{\textrm{peak}}^{\textrm{NR}}/2)^{-2/3}$.

\subsection{Spin-induced quadrupole-monopole terms}

\red{[check notations have been introduced]} For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. It includes consistently spin-orbit interactions up to 3.5PN order and spin-spin interactions up to 2PN order. In addition, the \texttt{SEOBNRv4T} model accounts for the spin-induced quadrupole moment of neutron stars~\cite{Poisson1998}, an extended-body effect distinguishing neutron stars from black holes. This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}. Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{}.}, at the moment the effect is only consistently included at the leading 2PN order. We leave for future work an extension to higher orders\footnote{\red{complete this} We note that TaylorF2 models do include all spin-square terms at 3PN, see App.~\ref{sec:taylorf2}. They are also present in the IMRPhenomPv2+NRTides model, but not in the SEOBNRv4ROM+NRTides model. TEOBResumS is also limited to the leading 2PN order.}.

>>>>

% tar -cvzf paper.tgz paper.tex paper.bbl hoff.pdf dhofs.pdf trainingset3d.pdf trainingset2d.pdf uncertaintysampling.pdf mismatch.pdf htildemaxerror.pdf hmaxerror.pdf Models_timings_FD_monotonic.pdf SEOBNRv4T_surrogate_speedup_deltaF_monontonic.pdf chirpmass.pdf q.pdf chieff.pdf tildelambda.pdf chieff_q.pdf

@@

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817. We present a frequency-domain surrogate version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}. This model includes the quadrupolar and octopolar adiabatic and dynamical tides. The version presented here is improved by the inclusion of the spin-induced quadrupole moment effect, and completed by a prescription for tapering the end of the waveform to qualitatively reproduce numerical relativity simulations. The resulting model has 14 intrinsic parameters. We reduce its dimensionality by using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. The implementation of the time-domain model can take up to an hour to evaluate using a starting frequency of 20~Hz, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations. We therefore construct a fast and faithful frequency-domain surrogate of this model using Gaussian process regression. The resulting surrogate has a maximum mismatch of $4.5\times 10^{-4}$ for the Advanced LIGO detector, and requires $\sim 0.13$~s to evaluate for a waveform with a starting frequency of 20~Hz. Finally, we perform an end-to-end test of the surrogate with a set of parameter estimation runs, and find that the surrogate accurately recovers the parameters of injected waveforms.

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can measure BNS properties such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}. Furthermore, an estimate of the BNS merger rate of 110--3840 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{O2Catalog} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be ``stacked'' to significantly improve measurements of the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.

The reliability of these measurements, however, depends on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data. Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Most time-domain waveform implementations involve solving a set of ordinary differential equations, which depending on the model can take seconds to hours for long BNS waveforms. The results then need to be transformed into the frequency domain where the analysis takes place. Because of this, analytic approximations providing waveforms directly in the frequency-domain  are often used instead.

For the analysis of GW170817 in Refs.~\cite{BNSPE, O2Catalog}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT), or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

In this paper, we will use an implementation of the EOB formalism that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018, NagarMessinaRettegno2018}, that uses a different point-particle and spin prescription. For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.

Although these EOB models are more expensive than frequency-domain models, parameter estimation results are still possible. In Refs.~\cite{BNSPE, O2Catalog}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. 
The measured tidal parameter using EOB waveforms and \texttt{RapidPE} were broadly consistent with the results using frequency-domain waveforms and traditional samplers (see Fig. 9 of Ref.~\cite{O2Catalog}). However, a direct comparison using identical waveforms has not yet been published for BNS systems.
As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for rapidly evaluating waveform models. The essential idea is to construct a fitting function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions. The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}. In this work, instead of using a reduced basis, we will instead use the fact that the Fourier-transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we can efficiently represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

@@

This work also takes a different approach from the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas}, multi-banding~\cite{VinciguerraVeitchMandel2017}, and relative binning~\cite{Zackay:2018qdy}.

We organize the paper as follows. In Section~\ref{sec:eob} we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In Section~\ref{sec:surrogate} we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section~\ref{sec:performance}. In Section~\ref{sec:pe} we verify that the surrogate can correctly extract the parameters of injected waveforms. Finally, we discuss future improvements in Section~\ref{sec:discussion} and give the expressions for the TaylorF2 base model in the Appendix.

@@

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. We adopt the tidally-augmented expression for $\Delta_u$ discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}:
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass EOB term (Eq.~(2.2) of Ref.~\cite{Bohe:2016gbl}), and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. If either component of the binary is a black hole, then we set the tidal polarizabilities to zero. The tidal contribution, including quadrupolar and octupolar dynamical tides, is
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,.
\end{align}
\end{widetext}
Here, $m_{A,B}$ are the masses of bodies $A$ and $B$, $M = m_A + m_B$ is the total mass, $X_{A,B}=m_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=m_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/m_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$. Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor~\cite{Steinhoff:2016rfi, Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.

Introducing a Keplerian orbital frequency defined from the radius variable as $M\Omega \equiv u^{3/2}$, using the notations (dropping the indices $A,B$)
\begin{subequations}
\begin{align}
	x = \frac{\hat{\omega}_{0\ell}}{m\Omega} \,, \; \epsilon \equiv \frac{256\eta}{5} \left( \frac{\hat{\omega}_{0\ell}}{m} \right)^{5/3} \,, \; \hat{t} \equiv \frac{8}{5\sqrt{\epsilon}} \left[ 1 - x^{5/3} \right] \,,
\end{align}
\end{subequations}
the dynamical enhancement factor  of~\cite{Steinhoff:2016rfi, Dietrich:2017feu} reads
\be\label{eq:kldyn}
	\hat{k}_{\ell, \rm{dyn}}(u; \omega_{0\ell}) = a_{\ell} + b_{\ell} \left[ f(x) +\sqrt{\frac{\pi}{3 \epsilon}} x^{2} \mathcal{Q} \right]\,,
\ee
with $a_{\ell}$, $b_{\ell}$ constant coefficients with relevant values $\{a_{2}, b_{2}\} = \{1/4, 3/4\}$ and $\{a_{3}, b_{3}\} = \{3/8, 5/8\}$. These expressions are used in practice for $\ell= m$. Here, we have rewritten the first two resonant terms of Eq.~(11) of~\cite{Dietrich:2017feu} as
\be\label{eq:deff}
	f(x) = x^{2} \left[ \frac{1}{x^{2} - 1} + \frac{5}{6} \frac{1}{1 - x^{5/3}} \right] \,.
\ee
Although the two terms are individually divergent\footnote{This requires to single out the near-resonance region in a numerical implementation.} at the resonance $x=1$, the function $f(x)$ is actually  regular, taking the value $f(1) = -1/12$. The last term in~\eqref{eq:kldyn} above reads
\begin{align}
	\mathcal{Q} &= \cos\left( \frac{3}{8} \hat{t}^{2} \right) \left[ 1 + 2 F_{S} \left( \frac{\sqrt{3}}{2\sqrt{\pi}} \hat{t} \right) \right] \nonumber\\
	& - \sin\left( \frac{3}{8} \hat{t}^{2} \right) \left[ 1 + 2 F_{C} \left( \frac{\sqrt{3}}{2\sqrt{\pi}} \hat{t} \right) \right] \,,
\end{align}
with $F_{S}$, $F_{C}$ the Fresnel sine and cosine functions. This term is regular at resonance $\hat{t} = 0$. The enhancement factor~\eqref{eq:kldyn} is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.

Tidal effects also enter the radiative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
\begin{equation}
h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
\end{equation} 
where the point-mass piece $h_{\ell m}^{\textrm{pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and the tidal piece $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.

The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the gravitational-wave flux (Eqs.~(5) and~(6) of Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived. The Hamiltonian equations of motion with this radiation-reaction force are numerically integrated beginning with the quasicircular initial conditions described in Ref.~\cite{Buonanno:2005xu}. The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics. We only include the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes when evaluating the radiation reaction force.

\subsection{Spin-induced quadrupole-monopole terms}
\label{subsec:QM}
